---
title       : "Data Analysis 4"
subtitle    : "BIOL10272: Practical Techniques"
author      : Dr Axel Barlow
job         : "email: axel.barlow@ntu.ac.uk"
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : zenburn      # {zenburn, tomorrow, solarized-dark, ...}
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {selfcontained, standalone, draft}
knit        : slidify::knit2slides
logo        : ntu-shield.png
biglogo     : NTU_open-graph.png
assets      : {assets: ../../assets}
license     : by-nc-sa
github:
  user: draxelbarlow
  repo: BIOL10272_data_analysis_4
  branch: "gh-pages"
---

```{r setup, include = FALSE}
library(tidyverse)
library(gganimate)
library(RColorBrewer)
library(knitr)
library(palmerpenguins)
```

<!-- adding bold and italic options -->
<style>
em {
  font-style: italic
}
strong {
  font-weight: bold;
}
</style>

## Data analysis 4

- Visualising two quantitative variables: scatterplots
- Dependent and independent variables
- Fitting a linear relationship: `least squares`
- Correlation coefficients: `r`
- Interpreting a correlation
- Quantifying the linear relationship
- Modelling the relationship: linear regression

--- .segue .dark 

## Visualising two quantitative variables

--- .class bg:white

## One categorical variable, one quantitative variable

```{r, out.width = "100%", out.height = "100%", fig.width = 12, fig.height = 7, dpi = 600, fig.align = "center", echo = FALSE}
gen_chi <- filter(penguins, species %in% c("Chinstrap", "Gentoo"))
gen_chi <- gen_chi %>% drop_na()

plot_bar2 <- ggplot(gen_chi, aes(species, body_mass_g, colour=species)) +
  geom_jitter(width=0.15, cex=3) +
  ylab("body mass (g)") +
  theme_bw(base_size = 25) +
  
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
  )

plot_bar2
```

--- .class bg:white

## Two quantitative variables

```{r, out.width = "95%", fig.width = 12, fig.height = 7, dpi = 600, fig.align = "left", echo = FALSE}
plot_scatter1 <- ggplot(gen_chi, aes(body_mass_g, flipper_length_mm)) +
  geom_point(cex=5, pch=21, fill="blue") +
  theme_bw(base_size = 25) +
  xlab("body mass (g)") +
  ylab("wing length (mm)") +
  
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
  )

plot_scatter1
```

--- .class #id

## Which variable goes on which axis?

- We need to identify the `independent` and `dependent` variables
- The independent variable goes on the *x* axis
- The dependent variable goes on the *y* axis

### In an experiment
- The independent variable is **the thing we manipulate**
- The dependent variable is **the thing we expect to change**
- We hypothesise the independent variable has an effect on the dependendent variable
- The null hypothesis is no effect

--- .class #id

## Examples

Dependent variable|Independent variable
---------------------|--------------------
Number of species|Habitat
Blood pressure|Drug test and control groups
Number of genes|Free living or parasitic microorganism
Disease prevalence|Country
Gene expression|Developmental stage

--- .class bg:white

## ELISA assay development

- Measure optical density at known protein concentrations

```{r, out.width = "85%", fig.width = 12, fig.height = 7, dpi = 600, fig.align = "left", echo = FALSE}
plot_scatter <- ggplot(DNase, aes(conc, density)) +
  geom_point(cex=5, pch=21, fill="blue") +
  theme_bw(base_size = 25) +
  xlab("Protein concentration (ng/mL)") +
  ylab("Optical density") +
  
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
  )

plot_scatter
```

--- .segue .dark 

## Sometimes, it's uncertain

--- .class bg:white

## Exploratory analysis of Gentoo penguin data

```{r, out.width = "95%", fig.width = 12, fig.height = 7, dpi = 600, fig.align = "left", echo = FALSE}
gen <- filter(penguins, species %in% c("Gentoo"))
gen <- gen %>% drop_na()

plot_scatter3 <- ggplot(gen, aes(bill_length_mm, bill_depth_mm)) +
  geom_point(cex=5, pch=21, fill="purple") +
  theme_bw(base_size = 25) +
  xlab("Bill length (mm)") +
  ylab("Bill depth (mm)") +
  
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
  )

plot_scatter3
```

--- .class bg:white

## Exploratory analysis of Gentoo penguin data

```{r, out.width = "95%", fig.width = 12, fig.height = 7, dpi = 600, fig.align = "left", echo = FALSE}
gen <- filter(penguins, species %in% c("Gentoo"))
gen <- gen %>% drop_na()

plot_scatter3 <- ggplot(gen, aes(bill_depth_mm, bill_length_mm)) +
  geom_point(cex=5, pch=21, fill="purple") +
  theme_bw(base_size = 25) +
  ylab("Bill length (mm)") +
  xlab("Bill depth (mm)") +
  
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
  )

plot_scatter3
```

--- .segue .dark 

## Fitting a linear relationship

--- .class #id

## Correlation

A connection, or relationship, between two or more things

### In Biosciences

A linear relationship between two quantitative variables

--- &twocol

## Linear relationship

- The points in the scatterplot form a straight line

*** =left

```{r, out.width = "90%", fig.width = 6, fig.height = 5, dpi = 600, fig.align = "left", echo = FALSE}
pos_eg <- data.frame(var1=c(1:10), var2=c(1:10))

plot_scatter1 <- ggplot(pos_eg, aes(var1, var2)) +
  geom_point(cex=4, pch=21, fill="coral") +
  theme_bw(base_size = 20) +
  xlab("Quantitative variable 1") +
  ylab("Quantitative variable 2")+
  
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
  )

plot_scatter1
```

- Positive correlation

*** =right

```{r, out.width = "90%", fig.width = 6, fig.height = 5, dpi = 600, fig.align = "left", echo = FALSE}
pos_eg <- data.frame(var1=c(1:10), var2=c(10:1))

plot_scatter2 <- ggplot(pos_eg, aes(var1, var2)) +
  geom_point(cex=4, pch=21, fill="coral") +
  theme_bw(base_size = 20) +
  xlab("Quantitative variable 1") +
  ylab("Quantitative variable 2")+
  
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
  )

plot_scatter2
```

- Negative correlation

--- .class bg:white

## Fitting a linear relationship to real data

- We can add a `line of best fit` using the method of `least squares`
- The straight line that minimises the square of the `residuals`

```{r, out.width = "80%", fig.width = 12, fig.height = 7, dpi = 600, fig.align = "center", echo = FALSE}
gen_chi <- filter(penguins, species %in% c("Chinstrap", "Gentoo"))
gen_chi <- gen_chi %>% drop_na()
fit <- lm(flipper_length_mm ~ body_mass_g, data=gen_chi)
gen_chi$predicted <- predict(fit)
gen_chi$residuals <- residuals(fit)

plot_scatter3 <- ggplot(gen_chi, aes(body_mass_g, flipper_length_mm)) +
  geom_point(cex=5, pch=21, fill="blue") +
  geom_smooth(method=lm, formula = y~x, se=FALSE, col="black", lwd=2) +
  theme_bw(base_size = 25) +
  xlab("body mass (g)") +
  ylab("wing length (mm)")+
  
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
  )

plot_scatter3
```

--- .class bg:white

## Fitting a linear relationship to real data

- We can add a `line of best fit` using the method of `least squares`
- The straight line that minimises the square of the `residuals`

```{r, out.width = "80%", fig.width = 12, fig.height = 7, dpi = 600, fig.align = "center", echo = FALSE}
plot_scatter3 +
   geom_segment(aes(xend = body_mass_g, yend = predicted), col="red")
```

--- .class #id

## Calculating the strength of the relationship

### If the residuals are small, the variables have a strong linear relationship

- We can measure this strength using a `correlation coefficient`
- We will use the `Pearson correlation coefficient`, which can be represented as `r`
- Scales from -1 to +1
- 0 = **no** linear relationship
- +1 = perfect **positive** linear relationship
- -1 = perfect **negative** linear relationship
- We can also obtain a `p-value`
- **Null hypothesis**: there is no linear relationship, the true `r` = 0

--- .class bg:white

## `r` shows the strength of correlation

```{r, out.width = "100%", fig.width = 12, fig.height = 6, dpi = 600, fig.align = "center", echo = FALSE}
#Title: An example of the correlation of x and y for various distributions of (x,y) pairs
#Tags: Mathematics; Statistics; Correlation
#Author: Denis Boigelot
#Packets needed : mvtnorm (rmvnorm), RSVGTipsDevice (devSVGTips)
#How to use: output()
#
#This is an translated version in R of an Matematica 6 code by Imagecreator.

library(mvtnorm)
library(RSVGTipsDevice)

MyPlot <- function(xy, xlim = c(-4, 4), ylim = c(-4, 4), eps = 1e-15) {
   title = round(cor(xy[,1], xy[,2]), 1)
   if (sd(xy[,2]) < eps) title = "" # corr. coeff. is undefined
   plot(xy, main = title, cex.main = 2, xlab = "", ylab = "",
        col = "darkblue", pch = 16, cex = 0.8,
        xaxt = "n", yaxt = "n", bty = "n",
        xlim = xlim, ylim = ylim)
}

MvNormal <- function(n = 1000, cor = 0.8) {
   for (i in cor) {
      sd = matrix(c(1, i, i, 1), ncol = 2)
      x = rmvnorm(n, c(0, 0), sd)
      MyPlot(x)
   }
}

rotation <- function(t, X) return(X %*% matrix(c(cos(t), sin(t), -sin(t), cos(t)), ncol = 2))

RotNormal <- function(n = 1000, t = pi/2) {
   sd = matrix(c(1, 1, 1, 1), ncol = 2)
   x = rmvnorm(n, c(0, 0), sd)
   for (i in t)
      MyPlot(rotation(i, x))
}

Others <- function(n = 1000) {
   x = runif(n, -1, 1)
   y = 4 * (x^2 - 1/2)^2 + runif(n, -1, 1)/3
   MyPlot(cbind(x,y), xlim = c(-1, 1), ylim = c(-1/3, 1+1/3))

   y = runif(n, -1, 1)
   xy = rotation(-pi/8, cbind(x,y))
   lim = sqrt(2+sqrt(2)) / sqrt(2)
   MyPlot(xy, xlim = c(-lim, lim), ylim = c(-lim, lim))

   xy = rotation(-pi/8, xy)
   MyPlot(xy, xlim = c(-sqrt(2), sqrt(2)), ylim = c(-sqrt(2), sqrt(2)))
   
   y = 2*x^2 + runif(n, -1, 1)
   MyPlot(cbind(x,y), xlim = c(-1, 1), ylim = c(-1, 3))

   y = (x^2 + runif(n, 0, 1/2)) * sample(seq(-1, 1, 2), n, replace = TRUE)
   MyPlot(cbind(x,y), xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5))

   y = cos(x*pi) + rnorm(n, 0, 1/8)
   x = sin(x*pi) + rnorm(n, 0, 1/8)
   MyPlot(cbind(x,y), xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5))

   xy1 = rmvnorm(n/4, c( 3,  3))
   xy2 = rmvnorm(n/4, c(-3,  3))
   xy3 = rmvnorm(n/4, c(-3, -3))
   xy4 = rmvnorm(n/4, c( 3, -3))
   MyPlot(rbind(xy1, xy2, xy3, xy4), xlim = c(-3-4, 3+4), ylim = c(-3-4, 3+4))
}

output <- function() {
#   devSVGTips(width = 7, height = 3.2) # remove first and last line for no svg exporting
#   par(mfrow = c(3, 7), oma = c(0,0,0,0), mar=c(2,2,2,0))
#   MvNormal(800, c(1.0, 0.8, 0.4, 0.0, -0.4, -0.8, -1.0));
#   RotNormal(200, c(0, pi/12, pi/6, pi/4, pi/2-pi/6, pi/2-pi/12, pi/2));
#   Others(800)
#   dev.off() # remove first and last line for no svg exporting

   par(mfrow = c(2, 4), oma = c(0,0,0,0), mar=c(2,2,2,0))
   MvNormal(800, c(1.0, 0.8, 0.4, 0.00));
   MvNormal(800, c(0.0, -0.4, -0.8, -1.0));

}

output()
```

*R code for plot by Denis Boigelot*

--- .class bg:white

## The slope does not affect `r`

```{r, out.width = "100%", fig.width = 12, fig.height = 6, dpi = 600, fig.align = "center", echo = FALSE}
   par(mfrow = c(2, 3), oma = c(0,0,0,0), mar=c(2,2,2,0))
   RotNormal(200, c(0, pi/12, pi/6));
   RotNormal(200, c(pi/2, pi/2-pi/12, pi/2-pi/6));
```

*R code for plot by Denis Boigelot*

--- .class bg:white

## Pearson correlation for penguin data

```{r, out.width = "95%", fig.width = 12, fig.height = 7, dpi = 600, fig.align = "left", echo = FALSE}
plot_scatter3 +
  ggtitle("r = 0.899, p < 0.01")

```

--- .segue .dark 

## Interpreting a correlation

--- .class #id

## Scientists have found the gene for...

<iframe src = 'https://news.sky.com/story/covid-19-gene-that-doubles-risk-of-death-from-covid-identified-by-oxford-university-scientists-12459684' height='600px'></iframe>

--- .class bg:white

## Eating ice cream does not cause dehydration

```{r, out.width = "90%", fig.width = 12, fig.height = 7, dpi = 600, fig.align = "center", echo = FALSE}
ice_can <- data.frame(icecream=jitter(1:10, factor=3), cancer=jitter(11:20, factor=5))

plot_scatter2 <- ggplot(ice_can, aes(icecream, cancer)) +
  geom_point(cex=5, pch=21, fill="darkred") +
  geom_smooth(method=lm, formula = y~x, se=FALSE, col="black", lwd=1) +
  theme_bw(base_size = 20) +
  xlab("Ice cream sales (thousand per month)") +
  ylab("Dehydration (% admissions per month)") +
  ggtitle("r = 0.977, p < 0.01")+
  
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
  )

plot_scatter2
```

--- .segue .dark 

## Quantifying the linear relationship

--- .class bg:white

## `r` does not provide information on slope

### but isn't the slope interesting?

```{r, out.width = "100%", fig.width = 12, fig.height = 4.5, dpi = 600, fig.align = "center", echo = FALSE}
library(mvtnorm)
library(RSVGTipsDevice)

MyPlot <- function(xy, xlim = c(-4, 4), ylim = c(-4, 4), eps = 1e-15) {
   title = round(cor(xy[,1], xy[,2]), 1)
   if (sd(xy[,2]) < eps) title = "" # corr. coeff. is undefined
   plot(xy, main = title, cex.main = 2, xlab = "", ylab = "",
        col = "darkblue", pch = 16, cex = 0.8,
        xaxt = "n", yaxt = "n", bty = "n",
        xlim = xlim, ylim = ylim)
}

MvNormal <- function(n = 1000, cor = 0.8) {
   for (i in cor) {
      sd = matrix(c(1, i, i, 1), ncol = 2)
      x = rmvnorm(n, c(0, 0), sd)
      MyPlot(x)
   }
}

rotation <- function(t, X) return(X %*% matrix(c(cos(t), sin(t), -sin(t), cos(t)), ncol = 2))

RotNormal <- function(n = 1000, t = pi/2) {
   sd = matrix(c(1, 1, 1, 1), ncol = 2)
   x = rmvnorm(n, c(0, 0), sd)
   for (i in t)
      MyPlot(rotation(i, x))
}

Others <- function(n = 1000) {
   x = runif(n, -1, 1)
   y = 4 * (x^2 - 1/2)^2 + runif(n, -1, 1)/3
   MyPlot(cbind(x,y), xlim = c(-1, 1), ylim = c(-1/3, 1+1/3))

   y = runif(n, -1, 1)
   xy = rotation(-pi/8, cbind(x,y))
   lim = sqrt(2+sqrt(2)) / sqrt(2)
   MyPlot(xy, xlim = c(-lim, lim), ylim = c(-lim, lim))

   xy = rotation(-pi/8, xy)
   MyPlot(xy, xlim = c(-sqrt(2), sqrt(2)), ylim = c(-sqrt(2), sqrt(2)))
   
   y = 2*x^2 + runif(n, -1, 1)
   MyPlot(cbind(x,y), xlim = c(-1, 1), ylim = c(-1, 3))

   y = (x^2 + runif(n, 0, 1/2)) * sample(seq(-1, 1, 2), n, replace = TRUE)
   MyPlot(cbind(x,y), xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5))

   y = cos(x*pi) + rnorm(n, 0, 1/8)
   x = sin(x*pi) + rnorm(n, 0, 1/8)
   MyPlot(cbind(x,y), xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5))

   xy1 = rmvnorm(n/4, c( 3,  3))
   xy2 = rmvnorm(n/4, c(-3,  3))
   xy3 = rmvnorm(n/4, c(-3, -3))
   xy4 = rmvnorm(n/4, c( 3, -3))
   MyPlot(rbind(xy1, xy2, xy3, xy4), xlim = c(-3-4, 3+4), ylim = c(-3-4, 3+4))
}

output <- function() {

   par(mfrow = c(1, 3), oma = c(0,0,0,0), mar=c(2,2,15,0))
   RotNormal(200, c(0, pi/12, pi/6));
}

output()
```

*R code for plot by Denis Boigelot*

--- .class #id

## Limitations of correlation

- We would like to be able to **quantify** how much a variable changes
- This would allow us to **predict** the value of one variable if we know the value of another
- Even `extrapolate` beyond the range of measurement

### To do this

- We assume that the 2 variables have a linear relationship (significant correlation)
- We set up a statistical `model` of how the relationship works
- We fit the data to the model using `linear regression`

--- .segue .dark 

## Modelling the relationship

--- &twocol bg:white

## Setting up the linear model

*** =left

- Our model has several `coefficients`
- These are like parts of the model:
  + The **slope** of the line
  + The **y-axis intercept**
- There is also variance, or error of the datapoints
- These are basically unknown and/or unsampled variables
- We can quantify how well the data is explained by the linear model using `R-squared`

*** =right

```{r, out.width = "80%", fig.width = 6, fig.height = 7, dpi = 600, fig.align = "center", echo = FALSE, message = FALSE}
#example <- data.frame(independent=c(1:10), dependent=jitter(11:20, factor=4))
example <- read_csv("./data/jitter_eg", col_names=TRUE)

plot_scatter2 <- ggplot(example, aes(independent, dependent)) +
  geom_point(cex=5, pch=21, fill="blue") +
  geom_smooth(method=lm, formula = y~x, se=FALSE, col="blue", lwd=1, fullrange=TRUE) +
  xlim(0, 10) +
  ylim(8, 21) +
  theme_bw(base_size = 20) +
  xlab("Independent variable") +
  ylab("Dependent variable") +
  geom_segment(aes(x = 2.1, xend = 7.5, y = 12, yend = 12), col="red") +
  geom_segment(aes(x = 7.5, xend = 7.5, y = 12, yend = 17.5), col="red")

plot_scatter2
```

--- &twocol bg:white

## Carrying out the regression

```{r, echo=FALSE}
mylm <- summary(lm(dependent ~ independent, data = example))
```

*** =left

- The regression fits the data to the linear model using least squares
- The slope is **`r format(round(mylm$coefficients[2], 3))`**
- The y-axis intercept is **`r format(round(mylm$coefficients[1], 3))`**
- R-squared is **`r format(round(mylm$adj.r.squared, 3))`**
- The remaining variation is due to unknown or unsampled variables

*** =right

```{r, out.width = "80%", fig.width = 6, fig.height = 7, dpi = 600, fig.align = "center", echo = FALSE}
plot_scatter2 
```

--- &twocol bg:white

## Now the really cool bit

*** =left

- We can predict the dependent variable for any value of the independent variable
- y = slope*x + intercept
- For example, x = 2.5
- y = 2.5*`r format(round(mylm$coefficients[2], 3))` + `r format(round(mylm$coefficients[1], 3))`
- y = `r (2.5 * mylm$coefficients[2]) + mylm$coefficients[1]`

**Now**

- Let's call slope *m*
- And the intercept *c*
- y = *mx* + *c*
- The general formula for a linear relationship

*** =right

```{r, out.width = "80%", fig.width = 6, fig.height = 7, dpi = 600, fig.align = "center", echo = FALSE}
plot_scatter2 
```

--- .segue .dark 

## The graph is the visualisation of the model

--- .class bg:white

## Linear regression for penguin data

Slope = 0.01384, y axis intercept = 0.01384, R-squared = 0.8063

```{r, out.width = "90%", fig.width = 12, fig.height = 7, dpi = 600, fig.align = "left", echo = FALSE}
gen_chi <- filter(penguins, species %in% c("Chinstrap", "Gentoo"))
gen_chi <- gen_chi %>% drop_na()
penlm <- summary(lm(flipper_length_mm ~ body_mass_g, data = gen_chi))

plot_scatter3 <- ggplot(gen_chi, aes(body_mass_g, flipper_length_mm)) +
  geom_point(cex=5, pch=21, fill="blue") +
  geom_smooth(method=lm, formula = y~x, se=FALSE, col="black", lwd=2) +
  theme_bw(base_size = 25) +
  xlab("body mass (g)") +
  ylab("wing length (mm)") +
  
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
  )

plot_scatter3
```

--- .class #id

## Data analysis 4

- Visualising two quantitative variables: scatterplots
- Dependent and independent variables
- Fitting a linear relationship: `least squares`
- Correlation coefficients: `r`
- Interpreting a correlation
- Quantifying the linear relationship
- Modelling the relationship: linear regression

--- &thankyou

## That's all folks!

**Next: revision session and exam**
